Gradual Soundness: Lessons from Static Python

### 2022-01-14 pj-7.1/paper.tex
pj: discussion draft

@@

\section{The Lessons}
\label{s:conclusion}

@@

%% TODO conclusions about SP? threats to validity?

\SP{} is ambitious in many ways:
\begin{itemize}
  \item Type checker and compiler for Python syntax
  \item Fork of Python 3.8
  \item Add significant runtime components: vtable, classloader
  \item Add a method-based JIT
  \item Aiming for tens of thousands of typed modules (surely not all 45k in the server, but otoh the plan must be to hit other parts of IG too)
\end{itemize}

\noindent{}But \SP{} is also carefully targeted:
\begin{itemize}
  \item
    Not aiming for an exhaustive type checker.
    The availability of tools like Pyre already do a fine job.
    (Don't forget that consequently, it's helpful for undefined types go to dynamic.)
  \item
    Shallow types come first in the implementation, then concrete / checked
    (ditto for adoption)
  \item
    Not aiming for the full gradual guarantees.
    Only really matters for the initial migration, the common case.
    Indeed, it's important to not have the guarantee for primitives for their performance.
  \item
    The type checker is module-at-a-time, not fine-grained.
\end{itemize}

\noindent{}Lesson: the migration works, with a dedicated support team.

%% 2022-01-06 TODO multiple dimensions of gradual

Sound types in Static Python give programmers a way to improve performance.
The change is not forced, not intrusive;
programmers can keep using unsound types when working toward a deadline.

Conjecture that soundness will find bugs too, and lead to more reliable products.
Too soon to say.

>>>>

%% 2022-01-14 TODO re 3.7%
%% That percentage represents the aggregate CPU improvement observed from all of
%% our individual diffs that converted modules to Static Python over the past
%% year. In a sizable number of cases a conversion to Static Python did mean that
%% functions in that module started being JITted that previously were not JITted,
%% because for most of the half we had a policy in place that all Static Python
%% modules are always JITted. (We’ve now decoupled these things for future
%% conversions, so opting functions into the JIT is now orthogonal to opting them
%% into SP.)
%% 
%% So the 3.7% does not include any CPU wins that Cinder or the JIT provide in any
%% non-static modules, it only includes improvements observed from specifically
%% converting a module from nonstatic to static. But it does in some cases include
%% the effects of going from nonstatic + no-JIT to SP+JIT for a given module. In
%% other cases a conversion may have gone from nonstatic+JIT to SP+JIT.
%% Unfortunately we don’t have any way to go back and tease apart these cases for
%% each individual module conversion.
%% 
%% What we are doing now is an experiment where we turn off SP in a set of hosts
%% and observe them relative to the rest of prod, without changing the JIT list.
%% This will show some % regression, which we expect to be some number <3.7%, and
%% that will give a solid figure for the benefit we are getting specifically from
%% SP across all static modules, totally independent of the JIT. Unfortunately
%% there have been some hiccups in this experiment and we don’t have that number
%% ready yet, but we hope to have it in the next week or so.

%% 2022-01-14 FILL TODO 
%% > I still have trouble understanding the CPU number. There must be many
%% > servers running many workloads and some statistics to merge them. Do
%% > you have a blog post or something that explains the profiling in more
%% > detail?
%% 
%% I don’t know of anything reasonably current in terms of blog posts. For IG web
%% prod, there are not “many workloads” — there is no specialization in the IG web
%% tier. There are many thousands of hosts, all running the exact same workload —
%% all those hosts are handling an undifferentiated random selection from a single
%% giant pool of incoming IG requests. So for all intents and purposes we can
%% consider them identical. (There are in practice some differences in the hosts
%% themselves — e.g. different generations of CPU hardware — that we do have to
%% take into account when running the experiments below.)
%% 
%% The details of how we measure a CPU regression or improvement are interesting,
%% though. What we do is we take a set of hosts that we want to measure (typically
%% an experiment group and a control group, where the control group is identical
%% to the rest of production and the experiment group has some change made to its
%% code or configuration whose effect we want to measure), and we slowly increase
%% the traffic to those hosts until they are at max CPU utilization — they cannot
%% handle any more traffic without falling over. Then we measure the
%% requests-per-second handled by each set of hosts; if the experiment group is
%% able to handle 2% more requests per second than the control group, while both
%% are at max CPU utilization, this effect is considered a 2% global CPU win.
%% 
%% This metric has the advantage that it is highly holistic and results-oriented;
%% it is a measure of our actual capacity to serve more requests with fewer hosts,
%% and thus it can subsume and account for, and give apples-to-apples comparison
%% of, many different possible reasons for an increase or decrease in a host’s
%% efficiency in serving requests.
%% 
%% The disadvantage of the metric is that the raw number it measures (requests per
%% second) is also highly dependent on the product code and client behavior at
%% that moment in time, and is not stable over time, because the average cost in
%% CPU cycles of “one request” is not stable over time. If the behavior of the IG
%% iOS or Android app changes so that it is sending more heavy requests and fewer
%% lightweight requests, RPS across the fleet will naturally decrease even though
%% the efficiency of the servers has not changed. Therefore, this metric is really
%% good for saying “at X point in time, Y change has a Z% overall effect on our
%% CPU efficiency” but it cannot be used to compare raw numbers from two different
%% points in time; we cannot use it to say “between X1 point in time and X2 point
%% in time, RPS increased Z%, therefore our efficiency has increased by Z%.”
%% 
%% So the 3.7% number we gave you is an aggregate win from doing many such
%% individual experiments over the course of the past year, where each experiment
%% was measuring the effect of converting X module (or small set of modules) from
%% nonstatic to static (where in some cases it was nonstatic+JIT to SP+JIT, and in
%% other cases it was nonstatic noJIT to SP+JIT.) And the updated number we hope
%% to get you soon is an overall measure from a single experiment, today, of the
%% effect of disabling SP everywhere, while leaving the JIT list untouched. We
%% expect the latter number to be smaller than 3.7% because it will remove the JIT
%% factor entirely, but there’s also an unknown factor (which could net positive
%% or negative) of how the behavior of long-ago-converted SP modules has changed,
%% or the capabilities of SP has changed, since their conversion, which changes
%% the net win of SP on those modules since the time when we converted them.

@@

\section{Lessons}
\label{s:conclusion}

@@

\SP{} is an ambitious undertaking.
Its developers are maintaining a fork of the Python 3.8 runtime,
a type checker,
an optimizing ahead-of-time compiler,
and a method-based JIT compiler.
The team has converted hundreds of untyped modules to \SP{}
and plans to convert thousands more in the future.

And yet, \SP{} is carefully restricted to set attainable goals
for its implementors and for adopters:
\begin{itemize}
  \item
    The syntax of sound static types is relatively small compared
    to Pyre~\cite{pyre} and other types-for-Python proposals~\cite{FILL}.
    Higher-order types are absent.
    The focus is instead on first-order class types, base types, and standard data structures.
  \item
    The type checker operates at a module-level granularity rather
    than at the level of statements or expressions.
  \item
    Gradual class hierarchies are permitted, but restricted to single inheritance.
    Furthermore, static-to-static method overrides cannot reduce the precision of types---a restriction
    that enables optimizations but breaks the static gradual guarantee.
  \item
    Shallow types such as $\sptrawpydict$ accommodate Python code but offer coarse
    soundness guarantees.
    These types have been the top priority for \SP{}.
    Concrete types support full soundness, but impose nonlocal maintenace costs.
    \SP{} supports few concrete types at the moment and plans to add more.
  \item
    Progressive primitive types are available to help developers maximize performance.
    These types are intentionally incompatible with the dynamic type to help programmers
    avoid casts at run-time.
\end{itemize}
%
Many of these restrictions go against the grain of the mainstream
research community, but they have been effective in practice.
The current \CPUchange{} increase in CPU usage is a huge improvement
at the scale of the Instagram web server.

In conclusion, we conjecture that the \SP{} approach to gradual typing
is a viable way to incrementally add types to a production codebase.
Determining whether this conjecture holds more broadly is a critical
question that calls for research on two fronts.
First, \SP{} must be applied to additional projects in a module-at-a-time manner.
A exhaustive performance evaluation~\cite{FILL} of a few small projects would be
especially useful.
Second, the \SP{} approach should be adapted to other languages.
An optionally-typed language such as TypeScript would be an ideal starting
point for such an experiment.
